---
title: "Dodgers vs. Yankees World Series Prediction"
author: "Gustave Miller"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Abstract

I was thinking about this upcoming world series and how we are watching 2 
powerhouse teams go at it for the world championship. Being a Dodgers fan, this 
got me thinking, what can I do to predict the outcomes of the games our nation 
will be following for the next week. So I will be first predicting the outcome 
of the first game which happens to just be a few hours away. From this, I will 
then be using a Bayesian model to further predict the outcome of the rest of 
the series until one team wins it all.

```{r}
# Load the datasets and necessary libraries:
library(dplyr)
library(knitr)

batters <- read.csv("C:/Users/gusmi/Dropbox (Your team)/PC/Desktop/Baseball Analytics/2024WS-Predict/DodgersYankees_Batter.csv")
pitchers <- read.csv("C:/Users/gusmi/Dropbox (Your team)/PC/Desktop/Baseball Analytics/2024WS-Predict/DodgersYankees_Pitcher.csv") 
```

## Data

Here is an example of what both the Pitcher dataset looks like as well as the
Batter dataset

```{r}
head(pitchers)
head(batters)
```

## Data Variables

The data I will be using for this project constitutes of both pitcher and 
batter data to allow me to produce the most accurate result I can. I will be 
using data from both teams, pitching and batting, from ONLY the 2024 postseason 
so far. I am doing this because regular season stats do not entirely account 
for a players performance in the postseason as we have seen so far. Thus I will 
be including the players who have at least played/pitched 1 game within this 
postseason and using that for the simulation.

Below you will find data dictionaries for both Batters and Pitchers discussing 
the stats I will be utilizing within my functions to simulate at-bats.

```{r include=FALSE}
#Data Variables that I will be using/looking at for my monte carlo simulation
# Define data for batters
batters_stats <- data.frame(
  Statistic = c("AVG", "OBP", "SLG", "OPS", "wOBA", "wRC+", "K%", "BB%"),
  Definition = c(
    "Batting Average - Ratio of hits to at-bats.",
    "On-Base Percentage - Ratio of times on base to total plate appearances.",
    "Slugging Percentage - Total bases per at-bat.",
    "On-base Plus Slugging - Sum of OBP and SLG.",
    "Weighted On-Base Average - Overall offensive contribution, adjusted for 
    run value.",
    "Weighted Runs Created Plus - Offensive performance metric, adjusted for 
    league average.",
    "Strikeout Percentage - Ratio of strikeouts to total plate appearances.",
    "Walk Percentage - Ratio of walks to total plate appearances."
  )
)

# Define data for pitchers
pitchers_stats <- data.frame(
  Statistic = c("ERA", "FIP", "WHIP", "K/9", "BB/9", "HR/9", "BABIP", "WAR"),
  Definition = c(
    "Earned Run Average - Average earned runs allowed per nine innings.",
    "Fielding Independent Pitching - Estimates a pitcher's run prevention 
    independent of fielding.",
    "Walks and Hits Per Inning Pitched - Measure of how many baserunners a 
    pitcher allows.",
    "Strikeouts per 9 Innings - Average strikeouts per nine innings pitched.",
    "Walks per 9 Innings - Average walks allowed per nine innings pitched.",
    "Home Runs per 9 Innings - Average home runs allowed per nine innings pitched.",
    "Batting Average on Balls in Play - Ratio of hits to balls in play.",
    "Wins Above Replacement - Overall value of a player compared to a 
    replacement-level player."
  )
)

```
```{r}
# Create tables using kable
cat("Batter Statistics") 
kable(batters_stats, col.names = c("Statistic", "Definition"))
```

```{r}
print("Pitcher Statistics")
kable(pitchers_stats, col.names = c("Statistic", "Definition"))
```


## Monte Carlo Simulation

First is the Monte Carlo Simulation I am using to simulate game 1 of the 2024
World Series. The starting pitchers for this game were released early so I was
able to account for that in my simulation by selecting them from the dataset. 
Since I completed this prior to the starting lineups being released, I will be
using 9 batters from the dataset from each team, there may be some error in 
this as there is not a guaruntee those I select to start will be starting. 

```{r}
# Filter to select the starting pitchers
dodgers_pitcher <- pitchers[pitchers$Name == 'Jack Flaherty' & 
                              pitchers$Team == 'LAD' & pitchers$Type == 'SP', ]
yankees_pitcher <- pitchers[pitchers$Name == 'Gerrit Cole' & 
                              pitchers$Team == 'NYY' & pitchers$Type == 'SP', ]

# Select 9 batters from each team to act as the starting 9
dodgers_batters <- batters[batters$Team == 'LAD', ][1:9, ]
yankees_batters <- batters[batters$Team == 'NYY', ][1:9, ]
```

### Function to simulate an at-bat of a batter

The code in this chunk ultimately simulates an at-bat of a batter using the
statistics present in the dataset I have chosen to use, this includes
calculating the odds of a specific type of hit as well as a walk or strikeout.

```{r}
# Function to simulate a single at-bat for each batter
simulate_at_bat <- function(batter, pitcher) {
  # Calculate probabilities for the at-bat
  k_prob <- pitcher$K.9 / 9
  bb_prob <- pitcher$BB.9 / 9
  hit_prob <- max(0, batter$AVG - (k_prob + bb_prob)) # Ensure non-negative
  iso_prob <- batter$ISO
  single_prob <- hit_prob * 0.5
  double_prob <- iso_prob * 0.3
  triple_prob <- iso_prob * 0.1
  hr_prob <- iso_prob * 0.1
  
  # Ensure that the probabilities sum to 1
  total_prob <- k_prob + bb_prob + single_prob + double_prob + triple_prob + hr_prob
  k_prob <- k_prob / total_prob
  bb_prob <- bb_prob / total_prob
  single_prob <- single_prob / total_prob
  double_prob <- double_prob / total_prob
  triple_prob <- triple_prob / total_prob
  hr_prob <- hr_prob / total_prob
  
  outcome <- sample(c('K', 'BB', '1B', '2B', '3B', 'HR'), size = 1, 
                    prob = c(k_prob, bb_prob, single_prob, double_prob, triple_prob, hr_prob))
  return(outcome)
}
```

### Function for updating the bases and score when there is a hit

This function here will help to update the bases after each batters outcome
if it is a hit, if its not a hit (Strikeout or Walk), that is taken care of
in the game simulation function. There is also some minor error here, as 
players can make errors on fielding causing base runners to progress further
or that there is a chance for a runner on second to score from a single. This
function will return a score value and the updated bases from the hit.

```{r}
# Function to update bases and score based on the hit type/outcome of the hit
update_bases <- function(bases, hit_type) {
  score <- 0
  if (hit_type == '1B') {
    score <- bases[3]
    bases <- c(1, bases[1:2])
  } else if (hit_type == '2B') {
    score <- bases[2] + bases[3]
    bases <- c(0, 1, bases[1])
  } else if (hit_type == '3B') {
    score <- bases[1] + bases[2] + bases[3]
    bases <- c(0, 0, 1)
  } else if (hit_type == 'HR') {
    score <- sum(bases) + 1
    bases <- c(0, 0, 0)
  }
  return(list(bases = bases, score = score))
}
```

### Inning Simulation function

This next function is to simulate a single inning up until 3 outs are recorded.
There is a lot occuring within this function such as keeping track of the score,
bases, outs, as well as determining if a relief pitcher should relieve the 
starting pitcher based on either amount of innings pitched, and amount of runs
given up. This function handels Walks and Strikeouts as well, and returns
the amount of runs scored in the inning.

```{r}
# Function to simulate an inning with potential pitcher change
simulate_inning <- function(batters, pitcher, relief_pitchers=NULL, inning=1,
                            runs=0) {
  outs <- 0
  score <- 0
  bases <- c(0, 0, 0) # 000: bases empty, 100: runner on first, etc.
  
  while (outs < 3) {
    batter <- batters[sample(nrow(batters), 1), ]
    
    # Check if a pitcher change is needed
    if (!is.null(relief_pitchers) && inning >=5 && runs >= 5 && 
        nrow(relief_pitchers) > 0) {
      pitcher_to_use <- relief_pitchers[sample(nrow(relief_pitchers), 1), ]
      relief_pitchers <- relief_pitchers[-1, ]
    } else {
      pitcher_to_use <- pitcher
    }
    
    outcome <- simulate_at_bat(batter, pitcher_to_use)
    
    if (outcome == 'K') {
      outs <- outs + 1
    } else if (outcome == 'BB') {
      if (bases[1] == 1) {
        if (bases[2] == 1) {
          if (bases[3] == 1) {
            score <- score + 1
            bases[3] <- 0
          } else {
            bases[3] <- 1
          }
        } else {
          bases[2] <- 1
        }
      } else {
        bases[1] <- 1
      }
    } else {
      result <- update_bases(bases, outcome)
      bases <- result$bases
      score <- score + result$score
    }
  }
  
  return(score)
}
```

### Game Simulation

This function is the most important of all, it runs the simulation of the 
entire game, and records the winner. It will run through a for loop for 9 
iterations representing the 9 innings a game goes through, recording each run
and taking into account pitching swaps for relief pitchers, also adding another
constraint for if the pitcher has pitched to at least 25 batter if they have
not given up the 5 runs or reached 5 innings since simulating amount of pitches
is not happening here.

```{r}
# Function to simulate a game with potential pitcher changes
simulate_game <- function(dodgers_batters, yankees_batters, dodgers_pitcher, 
                          yankees_pitcher) {
  dodgers_score <- 0
  yankees_score <- 0
  dodgers_batters_faced <- 0
  yankees_batters_faced <- 0
  
  # Initialize the full list of relief pitchers for each game simulation
  dodgers_relief_pitchers <- pitchers[pitchers$Team == 'LAD' & pitchers$Type == 'RP', ]
  yankees_relief_pitchers <- pitchers[pitchers$Team == 'NYY' & pitchers$Type == 'RP', ]
  
  # Initialize runs allowed trackers for each pitcher
  dodgers_pitcher_runs_allowed <- 0
  yankees_pitcher_runs_allowed <- 0
  
  for (inning in 1:9) {
    
    # Dodgers pitcher criteria
    if (inning >= 5 || (dodgers_pitcher_runs_allowed >= 5) || 
        dodgers_batters_faced >= 25) {
      if (nrow(dodgers_relief_pitchers) > 0) {
        selected_dodgers_relief <- dodgers_relief_pitchers[sample(nrow(dodgers_relief_pitchers), 1), ]
        inning_score <- simulate_inning(dodgers_batters, selected_dodgers_relief, dodgers_relief_pitchers, inning, dodgers_score)
        dodgers_score <- dodgers_score + inning_score
        dodgers_relief_pitchers <- dodgers_relief_pitchers[dodgers_relief_pitchers$Name != selected_dodgers_relief$Name, ]
      } else {
        inning_score <- simulate_inning(dodgers_batters, dodgers_pitcher, NULL, inning, dodgers_score)
        dodgers_score <- dodgers_score + inning_score
      }
    } else {
      inning_score <- simulate_inning(dodgers_batters, dodgers_pitcher, NULL, inning, dodgers_score)
      dodgers_score <- dodgers_score + inning_score
      dodgers_pitcher_runs_allowed <- dodgers_pitcher_runs_allowed + inning_score
      dodgers_batters_faced <- dodgers_batters_faced + 1
    }

    # Yankees pitcher criteria
    if (inning >= 5 || (yankees_pitcher_runs_allowed >= 5) || 
        yankees_batters_faced >= 25) {
      if (nrow(yankees_relief_pitchers) > 0) {
        selected_yankees_relief <- yankees_relief_pitchers[sample(nrow(yankees_relief_pitchers), 1), ]
        inning_score <- simulate_inning(yankees_batters,selected_yankees_relief,yankees_relief_pitchers, inning, yankees_score)
        yankees_score <- yankees_score + inning_score
        yankees_relief_pitchers <- yankees_relief_pitchers[yankees_relief_pitchers$Name != selected_yankees_relief$Name, ]
      } else {
        inning_score <- simulate_inning(yankees_batters, yankees_pitcher, NULL, inning, yankees_score)
        yankees_score <- yankees_score + inning_score
      }
    } else {
      inning_score <- simulate_inning(yankees_batters, yankees_pitcher, NULL, inning, yankees_score)
      yankees_score <- yankees_score + inning_score
      yankees_pitcher_runs_allowed <- yankees_pitcher_runs_allowed + inning_score
      yankees_batters_faced <- yankees_batters_faced + 1
    }
  }
  
  if (dodgers_score > yankees_score) {
    return('Dodgers')
  } else {
    return('Yankees')
  }
}
```


### Running the simulation

Now I am running the Monte Carlo Simulation with 10,000 simulations for this 
one game so I can return the most accurate result possible. Then takinig the 
mean amount of games each team won and returning that to see which team won
more. The output will have both teams probabilities of winnig this game 1 of
the series, however for the sake of this, the team with the higher probability
will be declared the winner of this game.

```{r}
set.seed(81)
# Run the Monte Carlo simulation with 10000 simulations
num_simulations <- 10000
results <- c()

# For reproducibility
for (i in 1:num_simulations) {
  winner <- simulate_game(dodgers_batters, yankees_batters, dodgers_pitcher, yankees_pitcher)
  results <- c(results, winner)
}

# Calculate win probabilities
dodgers_win_prob <- mean(results == 'Dodgers')
yankees_win_prob <- mean(results == 'Yankees')

cat(sprintf("Probability of Dodgers winning Game 1: %.2f%%\n", dodgers_win_prob * 100))
cat(sprintf("Probability of Yankees winning Game 1: %.2f%%\n", yankees_win_prob * 100))

if (dodgers_win_prob > yankees_win_prob){
  cat(sprintf("Dodgers favored to win game 1 with a probability of %.2f%%\n", dodgers_win_prob * 100))
} else {
    cat(sprintf("Yankees favored to win game 1 with a probability of %.2f%%\n", yankees_win_prob * 100))
}
```



# Bayesian Simulation to Determine the Result of Entire Series

This next simulation I am doing involves a Bayesian model to simulate the rest 
of the World Series between the Dodgers and Yankees. The simulation code 
utilizes starting pitchers' data and prior win probabilities from a the Monte 
Carlo simulation to calculate each team’s chances based on game results. 
Specifically, it applies a Bayesian update function, (bayesian_series_update), 
which recalculates probabilities after each game by factoring in the wins 
needed for each team to secure the series.

For each game, the simulation chooses a starting pitcher for each team, then 
runs the simulation based on the stats to then update the amount of wins for
the series each team has. Then outputs the winners of each game and the 
entire series crowning one team the World Series Champion.

### Obtaining Starting Pitchers

```{r}
# Function to get the starting pitcher for a specific game
get_pitcher_for_game <- function(game_number, team, pitcher_data) {
  if (team == "Dodgers") {
    return(pitcher_data %>% filter(Team == 'LAD' & Type == 'SP') %>% 
             slice(game_number))
  } else {
    return(pitcher_data %>% filter(Team == 'NYY' & Type == 'SP') %>% 
             slice(game_number))
  }
}
```

### Creating the bayesian function to update win probabilities

```{r}

# Updated bayesian update function for series win probability
bayesian_series_update <- function(prior, dodgers_wins, yankees_wins) {
  # Calculate the wins remaining for each team to win the series
  games_remaining_dodgers <- max(0, 4 - dodgers_wins)
  games_remaining_yankees <- max(0, 4 - yankees_wins)
  
  # Adjust probabilities based on remaining games needed
  prob_dodgers_win_series <- prior[1] * (1 / (games_remaining_dodgers + 1))
  prob_yankees_win_series <- prior[2] * (1 / (games_remaining_yankees + 1))
  
  # Normalize to ensure the probabilities sum to 1
  total_prob <- prob_dodgers_win_series + prob_yankees_win_series
  posterior <- c(prob_dodgers_win_series / total_prob, 
                 prob_yankees_win_series / total_prob)
  
  return(posterior)
}

```

### Function to simulate the series 

```{r}

simulate_series <- function(num_games, initial_probs, pitchers, game_1_winner) {
  # Initialize win counts
  dodgers_wins <- 0
  yankees_wins <- 0
  current_probs <- initial_probs
  
  # Loop through each game
  for (game in 1:num_games) {
    # Ensure that current_probs has exactly two elements
    if (length(current_probs) != 2) {
      stop("current_probs must contain exactly two elements.")
    }
    
    # Simulate game result based on current probabilities
    game_result <- sample(c("Dodgers", "Yankees"), size = 1, replace = TRUE, prob = current_probs)
    
    # Update win counts
    if (game_result == "Dodgers") {
      dodgers_wins <- dodgers_wins + 1
    } else {
      yankees_wins <- yankees_wins + 1
    }
    
    # Print game result
    cat("Game", game, "Winner:", game_result, "| \n")
    
    # Check if a team has won the series
    if (dodgers_wins == 4 || yankees_wins == 4) {
      break
    }
    
    # Adjust probabilities based on winner of the current game
    current_probs <- if (game_result == "Dodgers") {
      c(0.6, 0.4)  # Example adjusted probabilities if Dodgers win
    } else {
      c(0.4, 0.6)  # Example adjusted probabilities if Yankees win
    }
  }
  
  # Return final result of series
  return(list(final_probs = current_probs, dodgers_wins = dodgers_wins, yankees_wins = yankees_wins))
}
```

### Outputs and running of Simulation

```{r}
set.seed(81)
# Initial win probabilities from Monte Carlo simulation
initial_probs <- c(dodgers_win_prob, yankees_win_prob)

# Determine Monte Carlo result for Game 1
game_1_winner <- ifelse(dodgers_win_prob > yankees_win_prob, "Dodgers", "Yankees")

# Run series simulation
num_games <- 7
series_result <- simulate_series(num_games, initial_probs, pitchers, game_1_winner)

# Output final series result
cat(sprintf("\nFinal Series Results: Dodgers Wins = %d, Yankees Wins = %d\n", series_result$dodgers_wins, series_result$yankees_wins))

```
From these results here we can see that the Dodgers will be crowned the 
2024 World Series Champion.

## Biases and Flaws

There are a few biases present in this simulation as well as flaws that come
with it, such as the application of randomness when choosing relief pitchers, 
assuming the starting 9 batters. These are 2 examples which can change the 
outcome of a game depending on who is playing. Next I dont account for many 
other scenarios such as a batter being hit by a pitch, stealing a base, fielder
errors, a runners speed on the bases, the distance of the outfield fence for 
home runs, the types of pitches a pitcher uses. All of these and much more 
contribute to the outcome of a game and which I am not using in this model.

## Conclusion

The application of a Monte Carlo simulation, paired with an application of a 
Bayesian model, can help me to predict the outcome of this 2024 World Series
matchup between the Dodgers and Yankees. Now I can only wait and see how 
accurate my prediction is based on the actual outcome. 

## Citations

I used this site's search tool for both batting and pitching for both teams
Major League Leaderboards - 2024 - Pitching | FanGraphs Baseball. (n.d.). 
FanGraphs Baseball.
[site](https://www.fangraphs.com/leaders/major-league?stats=pit&lg=all&qual=0&type=8&season=2024&season1=2024&ind=0&rost=&age=&filter=&players=0&team=22&pos=all&pageitems=30&postseason=Y)

